Titeln på avhandlingens språk

### 2023-02-04 body/part1/background.tex
great

\medskip
Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
The conceptual foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
As an answer to the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
\citeauthor{satyanarayanan2009case} named these compute nodes \emph{cloudlets}, and envisioned them as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user (see \cref{fig:edgecomputing})
This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure.
This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.

The architecture of edge computing offers several advantages, the primary being the significant reduction of latency.
Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
Other advantages include bandwidth reduction and improved privacy.
By processing and aggregating data closer to where its needed, edge computing can drastically reduce the amount of traffic going to cloud datacenters.
Cloudlets also present opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.

In later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \acl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
These applications had been hitherto inviable to implement due to the limitations of cloud offloading, but have now become a practical reality thanks to \gls{MEC}.
We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.

\todo[inline]{Something feels missing here}

Finally, we must note a closely related concept; \emph{fog} computing.
Introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, fog computing concerns itself with the distribution of computing power between the cloud and the edge.
This stands in contrast to the stated primary goal of edge computing, which is the reduction of latency and jitter by offloading computation to infrastructure as close to the user as possible.
The motivating factor for the fog paradigm is the aggregation of data and reduction of bandwidth requirements for the \gls{IOT}.
By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive sensor networks~\cite{yi2015survey}.

@@

There is increasing interest from academia and industry in a novel category of wearable and mobile immersive applications called \gls{WCA}~\cite{ha2014towards,chen2018application,chen2015early,belletier2021wearable,chen2017empirical,wang2020scaling,chatzopoulos2017hyperion}
They represent a particular class of \gls{AR} applications, aiming to amplify human cognition in both day-to-day tasks and professional settings.
\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
These systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
More recently, their scope has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards,ha2014towards}.
Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}, and it is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.
\todo[inline]{Develop a bit further. Talk about more reactive setups like ping pong too.}

@@

We identify a number of defining characteristics of \gls{WCA}.
One, and as the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
they are pervasive and mobile~\cite{ha2014towards}.

Next, interaction with \glspl{WCA} is immersive and seamless.
As mentioned above, \glspl{WCA} operate in a manner similar to \gls{GPS} guidance systems.
Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
The input is then parsed to an internal symbolic representation of the state of the task, and whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
Feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
If no change in state is detected, the application remains silent and out-of-the-way of the user;
in other words, samples which capture an intermediate or unfinished state, or simply noise are silently discarded.
\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
Assistants should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
\gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.

Note that this mode interaction implies seamless integration of the application with the context of the user and the real world.
The user does not trigger an update explicitly at any point, leading to one of the key characteristics of \gls{WCA}.
The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
An important corollary of this is that these are therefore the \emph{only} points in time at which users can notice changes in system responsiveness.

Finally, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand) to the user.
This goes hand in hand with the previous characteristic of seamless interaction, as users will have expectations of constant and immediate feedback as they interact with the system.
As is the case for most \gls{XR} applications, delayed feedback in \gls{WCA} can have severe negative consequences for the quality of the user experience.
Depending on the task, these consequences can range from simply distracting or annoying the user (in the case of less interactive tasks such as step-by-step assembly), to actively handicapping user performance (in the case of highly interactive tasks).

The above characteristics together lead to interesting consequences for \gls{WCA} system design.
Their wearable nature directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
The requirement of immersive and seamless interaction suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
This kind of compute- and energy-intensive processing is unfeasible to be performed on wearable devices, and thus these applications necessarily make use of compute offloading.
As explained above, the cloud is unsuitable as an offloading point for these applications, given the high latencies incurred, and there is thus a broad understanding today that with the advent of edge computing, \glspl{WCA} will become a practical reality~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.

\medskip
In literature, research on \gls{WCA} has focused mostly on the implementation of prototype applications and platforms.
\citeauthor{ha2014towards}~\cite{ha2014towards} coined the term \acl{WCA} in\ \citeyear{ha2014towards}.
Their work described a first prototype system employing a Google Glass wearable device, and identified the challenges of offloading the computation for such an application to the cloud.
\cite{chen2015early} and~\cite{chen2018application} further develop this platform, and produce a number of different assistive applications such as the aforementioned LEGO and IKEA assistants, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis.
Yet another platform for \gls{WCA} is introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
Furthermore, important work has been done in the implementation of such systems for industrial and factory applications~\cite{belletier2021wearable}.

In later years, an body of work studying the optimization of these applications has emerged.
A seminal work in this field is~\cite{chen2017empirical}, in which the authors establish hard and soft latency bounds for acceptable performance in \gls{WCA}.
\citeauthor{wang2019towards}~\cite{wang2019towards} explore strategies for scalable \gls{WCA} on the edge through different adaptive sampling and resource allocation schemes;
this is further developed in the authors' PhD thesis later on~\cite{wang2020scaling}.
\cite{moothedath2021energy},\ \cite{moothedath2022energy1}, and~\cite{moothedath2022energy2} studied the potential for energy optimization in \gls{WCA} through convex optimization of the sampling strategies employed.
In~\cite{george2020openrtist}, the authors introduce \emph{OpenRTiST}, a benchmarking tool for immersive, latency-sensitive, and bandwidth-hungry applications on edge computing leveraging \glspl{DNN} to perform real-time style transfer on a live video feed.

The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
There's more to end-to-end latency, \gls{WCA} performance, and scalability than merely the question of where the compute backend is placed, however.
The design of any complex application such as \gls{WCA} involves a multitude of decisions with the potential to influence the system responsiveness as experienced by the user.
These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.add
Our work in~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} introduces the first ever model for realistic emulation of these applications on real hardware.
Our characterization of human timings in these systems in~\cite{olguinmunoz2021impact}, and our subsequent design of a realistic model for human behavior in \gls{WCA} constitute novel insights and tools for the optimization of these systems.

@@

\glsreset{TTF}%
From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
\begin{inlineenum}
    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
\end{inlineenum}
These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
We will refer to them extensively in \cref{chap:contributions}.

>>>>

\medskip
Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
The foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
To tackle the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
\citeauthor{satyanarayanan2009case} named these compute nodes \emph{cloudlets}, and envisioned them as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user (see \cref{fig:edgecomputing})
This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the cloud).
This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.

The architecture of edge computing offers several advantages, the primary being the significant reduction of latency.
Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
Another advantage is bandwidth reduction.
%By processing and aggregating data closer to where its needed, edge computing can significantly reduce the amount of traffic going to cloud datacenters.
A variant of edge computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for this purpose.
By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
Edge computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.

Finally, in later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \aclp{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
These applications had been hitherto inviable to implement due to the limitations of cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.

@@

There is increasing interest from academia and industry in a novel category of wearable and mobile immersive applications called \gls{WCA}~\cite{ha2014towards,chen2018application,chen2015early,belletier2021wearable,chen2017empirical,wang2020scaling,chatzopoulos2017hyperion}
They represent a particular class of \gls{AR} applications, aiming to amplify human cognition in both day-to-day tasks and professional settings.

@@

\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
These applications operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
The input is parsed to an internal symbolic representation of the state of the task;
whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
This feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
If no change in state is detected, the application remains silent and out-of-the-way.
In other words, the application silently discards samples which capture an intermediate or unfinished state, or simply are too noisy to be processed.
\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.

As the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
they are pervasive and mobile~\cite{ha2014towards}.
This leads to interesting consequences for \gls{WCA} system design when combined with their requirement of seamless integration with the context of the user.
The wearable nature of \gls{WCA} directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
The requirements of immersive and seamless interaction, on the other hand, suggest a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
This kind of compute- and energy-intensive processing, often involving \gls{ML} algorithms and \glspl{DNN}, is unfeasible to be performed on wearable devices, and thus these applications \emph{must} necessarily make use of compute offloading~\cite{ha2014towards,wang2020scaling}.

However, it is understood today the cloud represents an unsuitable candidate for offloading computation in these applications, given the high latencies involved.
Given their immersive nature, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand), as users will have expectations of constant and immediate feedback as they interact with the system.
As is the case for most \gls{XR} applications, delayed feedback in \gls{WCA} can have severe negative consequences for the quality of the user experience.
Depending on the task, these consequences can range from simply distracting or annoying the user (in the case of less interactive tasks such as step-by-step assembly), to actively handicapping user performance (in the case of highly interactive tasks).
There is therefore a broad understanding today that edge computing is a key enabling technology for \gls{WCA} applications~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.

\medskip
In literature, research on \gls{WCA} has focused mostly on the implementation of prototype applications and platforms.
\citeauthor{ha2014towards}~\cite{ha2014towards} coined the term \acl{WCA} in\ \citeyear{ha2014towards}.
Their work described a first prototype system employing a Google Glass wearable device, and identified the challenges of offloading the computation for such an application to the cloud.
The authors were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
More recently, the scope of \glspl{WCA} has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards}.
\cite{chen2015early} and~\cite{chen2018application} further developed the platform described in~\cite{ha2014towards}, and produced a number of different assistive applications such as the aforementioned LEGO assistant, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis, and more advanced assembly assistant for IKEA furniture.
Yet another platform for \gls{WCA} was introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
Furthermore, important work has been done in the implementation of such systems for industrial and factory applications~\cite{belletier2021wearable}.
Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
It is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.

In later years, a body of work studying the optimization of these applications has emerged.
A seminal work in this field is~\cite{chen2017empirical}, in which the authors establish hard and soft latency bounds for acceptable performance in \gls{WCA}.
\citeauthor{wang2019towards}~\cite{wang2019towards} explore strategies for scalable \gls{WCA} on the edge through different adaptive sampling and resource allocation schemes;
this is further developed in the authors' PhD thesis later on~\cite{wang2020scaling}.
\cite{moothedath2021energy},\ \cite{moothedath2022energy1}, and~\cite{moothedath2022energy2} studied the potential for energy optimization in \gls{WCA} through convex optimization of the sampling strategies employed.
In~\cite{george2020openrtist}, the authors introduce \emph{OpenRTiST}, a benchmarking tool for immersive, latency-sensitive, and bandwidth-hungry applications on edge computing leveraging \glspl{DNN} to perform real-time style transfer on a live video feed.

The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
There's more to end-to-end latency, \gls{WCA} performance, and scalability than merely the question of where the compute backend is placed, however.
The design of any complex application such as \gls{WCA} involves a multitude of decisions with the potential to influence the system responsiveness as experienced by the user.
These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.
On the other hand, recently published models for end-to-end latency of edge computing architectures, are quite complex, while not accounting for the specifics of \gls{WCA} and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite}.
Our work in~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} introduces the first ever model for realistic emulation of these applications on real hardware.
Our characterization of human timings in these systems in~\cite{olguinmunoz2021impact}, and our subsequent design of a realistic model for human behavior in \gls{WCA} constitute novel insights and tools for the optimization of these systems.

@@

\glsreset{TTF}%
From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
\begin{inlineenum}
    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
\end{inlineenum}
These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
We will refer to them extensively in \cref{chap:contributions}.
